```{r, echo=FALSE, purl=FALSE, message=FALSE}
knitr::opts_chunk$set(results='hide', comment = "#>", purl = FALSE)
```

# Domain specific graphs {#domains}

> Learning Objectives
>
> * Be aware of specialized graph packages and know where to look for them
> * Understand the basic structure of iheatmapr, tmap, and visNetwork examples
> * Modify parameters of provided graph examples

------------

`ggplot` can get you a long way, but if you need to do a particular, more complex graph it is worth checking if there might be an R package for that. Typically it would do one type of visualization and do that really well. Below are a few examples.

## Heatmaps (e.g. **`iheatmapr`**)

```{r heatmap-demo, results='show'}
library(iheatmapr)
data(measles, package = "iheatmapr")

main_heatmap(measles, name = "Measles<br>Cases", x_categorical = FALSE,
             layout = list(font = list(size = 8))) %>%
  add_col_groups(ifelse(1930:2001 < 1961,"No","Yes"),
                  side = "bottom", name = "Vaccine<br>Introduced?",
                  title = "Vaccine?",
                  colors = c("lightgray","blue")) %>%
  add_col_labels(ticktext = seq(1930,2000,10),font = list(size = 8)) %>%
  add_row_labels(size = 0.3,font = list(size = 6)) %>% 
  add_col_summary(layout = list(title = "Average<br>across<br>states"),
                  yname = "summary")  %>%                 
  add_col_title("Measles Cases from 1930 to 2001", side= "top") %>%
  add_row_summary(groups = TRUE, 
                  type = "bar",
                  layout = list(title = "Average<br>per<br>year",
                                font = list(size = 8)))
```



## Networks (e.g. **`visNetwork`**)

```{r network-demo, results='show'}
library('visNetwork') 
nodes <- read.csv("https://github.com/cengel/R-data-viz/raw/master/demo-data/network/Dataset1-Media-Example-EDGES.csv", header=T, as.is=T)
links <- read.csv("https://github.com/cengel/R-data-viz/raw/master/demo-data/network/Dataset1-Media-Example-EDGES.csv", header=T, as.is=T)
nodes$shape <- "dot"  
nodes$shadow <- TRUE # Nodes will drop shadow
nodes$title <- nodes$media # Text on click
nodes$label <- nodes$type.label # Node label
nodes$size <- nodes$audience.size # Node size
nodes$borderWidth <- 2 # Node border width

nodes$color.background <- c("slategrey", "tomato", "gold")[nodes$media.type]
nodes$color.border <- "black"
nodes$color.highlight.background <- "orange"
nodes$color.highlight.border <- "darkred"

visNetwork(nodes, links) %>%
  visOptions(highlightNearest = TRUE, 
             selectedBy = "type.label")
```


## Maps (e.g. **`tmap`**) 

For this demo we first need to download the data. If you have your working directory set to `R-data-viz` and it contains a folder called `data`, this will download and extract the map data into a subfolder of your `data` folder, called `map`:

```{r tmap-download, eval=FALSE}
download.file("https://github.com/cengel/R-data-viz/raw/master/demo-data/map.zip",
              "data/map.zip")
unzip("data/map.zip", exdir="data")
```

Now, here is the map.

```{r tmap-demo}
library(tmap)
library(tmaptools)
suppressPackageStartupMessages(library(sf))
              
london <- st_read("data/map", "London", quiet = TRUE)
crime_densities <- st_read("data/map", "Crimes", quiet = TRUE)
thames <- st_read("data/map", "Thames", quiet = TRUE)

tm_shape(crime_densities) +
  tm_fill(col = "level", palette = "YlOrRd", 
    title = expression("Crimes per " * km^2)) + 
tm_shape(london) + tm_borders() +
tm_shape(thames) + tm_lines(col = "steelblue", lwd = 4) +
tm_compass(position = c("left", "bottom")) +
tm_scale_bar(position = c("left", "bottom")) + 
tm_style_gray(title = "Crimes in Greater London\nOctober 2015")
```


Good starting places to look for additional examples are the R Graph Gallery: https://www.r-graph-gallery.com/all-graphs/ and the CRAN Task View: https://CRAN.R-project.org/view=Graphics. 
